Dark matter halo concentrations with a bayesian approach

### 2014-09-30 paper/article.tex
massive changes
Massive changes part II
anadiendo texto

\section{Introduction}
\label{sec:introduction}

@@

The most interesting feature of Eq. (\ref{eq:profile}) is that the
concentration is the only free parameter to describe the shape of the
halo densitry profile. In Figure \ref{fig:profiles} we show $m(<x)$ as
a function of $x$ for different values of the concentration in the range
$1\leq c \leq 20$.

@@

\section{A Bayesian Approach to Halo Fitting}
\label{sec:method}

We proceed to find the value of the concentration parameter that
best describes the simulation data, following the model in
Eq. (\ref{eq:profile}). We use the integrated mass profile because it
allows us to use the data directly with the simulation without binning
the particle positions and estimating a density.

We construct the integrated mass profile by ranking the particles by
their increasing distance to the center of the halo. Once they are ranked,
the total mass at a radius $r_i$, increases by $m_p$, where $r_i$ is
the position of the $i$-th particle and $m_p$ is the mass of the
computational particle.  We define the center of the halo to be  at
the position of the particle with the lowest gravitational
potential. In the process of building the mass profile we discard the
particle at the center.

We stop the construction of the integrated mass profile once we arrive
at an average density of $\Delta_h\bar{\rho}$. This radius marks the
virial radius and the virial mass. We divide the total mass enclosed
mass $M_i$ and the radii $r_i$ by these values to obtaine the
dimensionless variables $x_i$ and $m_i$.

Using these positions and masses we define the following $\chi^2$ function

\begin{equation}
\chi^2(c) = \sum_{i}[\log m_i - \log m(< x_i;c)]^2, 
\end{equation}
%
where $m(<x_i;c)$ corresponds to the values in Eq.(\ref{eq:profile}) at
$x=x_i$ and a given value of the concentration parameter $c$.

Finally we use a Metropolis-Hastings algorithm to sample the likelihood
function distribution defined by ${\cal L}(c)=\exp(-\chi^2(c)/2)$ to
find the optimum value of $c$ and its associated uncertainty
$\sigma_c$.

\section{Numerical Experiments}

@@

We generate $100$ mock halos with concentration values randomly
placed in the range $1<c<20$. Each one of the halos is generated with
four different total particle numbers: $20$, $200$, $2000$ and $20000$.

The method we use to generate the halos is based on the integrated
mass profile. Given the number of particles $n$ and the concentration
$c$ we define the mass element as $\delta m = 1/n$ (this corresponds
to the mass of each particle such that the total mass is one). Then
for each number $k$ from $1$ to $n$ we find the value of $r$ such that
the difference 
%
\begin{equation}
m(<r;c)-k \cdot \delta m 
\end{equation}
%
is zero using Ridders' method. This value of $r$ is the radius of
the $k$th particle of the generated halo, then polar and azimuthal
angles $\theta$ and $\phi$ are randomly generated. Finally these three
coordinates are transformed into cartesian coordinates
$(r,\theta,\phi) \rightarrow (x,y,z)$. This process is repeated $n$
times in order to generate the $x,y,z$ coordinates for each particle.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.48\textwidth]{error.pdf}
\end{center}
\caption{Relative error Against number of Particles
    \label{fig:error}}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
  \includegraphics[width=0.48\textwidth]{concentration.pdf}
\end{center}
\caption{Concentration against number of particles
    \label{fig:concentrations}}
\end{figure}

>>>>

\section{Introduction}
\label{sec:introduction}
In the context of the concordance cosmology, the matter contents of
the Universe are dominated by an unknown dark matter which behaves as
a collisionless fluid only perturbed by gravity.

Numerical experiments are able to follow the non-linear evolution of
dark matter down to galactic scales. On these scales a common
approximation considers the matter distribution as spherical with a
radial depedence. The function describing this distribution is known
as a radial density profile.

The consensus is that density profiles are almost universal;
almost independet of cosmological parameters and similar for different
objects after some simple physical re-scalig is applied.

The detailed halo structure influences the hosted galaxies.

The concentration parameter is imporant in galaxy formation studies.

Different studies vary about the result of the concentration-mass
relationship.

We propose a new way to obtain the concentration for halos in N-body
simulations.

We use the integrated mass profile because it
allows us to use the data directly with the simulation without binning
the particle positions and estimating a density.

As an application we study the mass concentration relationship.

@@

The most interesting feature of Eq. (\ref{eq:profile}) is that the
concentration is the only free parameter to describe the density
profile. In Figure \ref{fig:profiles} we show $m(<x)$ as 
a function of $x$ for different values of the concentration in the range
$1\leq c \leq 20$.

@@

In Figure XXX we show the circular velocity profile for the same
concentrations as in Figure \ref{fig:profiles}.

\section{A new Approach to Halo Fitting}
\label{sec:method}

We estimate the concentration parameter by fitting the
integrated mass profile described in Equation
(\ref{eq:profile}). First we define the center of the halo to be  at
the position of the particle with the lowest gravitational
potential.

We construct the integrated mass profile by ranking the particles by
their increasing distance to the center of the halo. Once they are ranked,
the total mass at a radius $r_i$, increases by $m_p$, where $r_i$ is
the position of the $i$-th particle and $m_p$ is the mass of the
computational particle.  In this process we discard the
particle at the center.

We stop the construction of the integrated mass profile once we arrive
at an average density of $\Delta_h\bar{\rho}$, with $\Delta_h=740$,
roughly corresponding to 200 times the critical density. This radius marks the
virial radius and the virial mass. We divide the total mass enclosed
mass $M_i$ and the radii $r_i$ by these values to obtain the
dimensionless variables $x_i$ and $m_i$.

Using these new variables we define the following $\chi^2$ function

\begin{equation}
\chi^2(c) = \sum_{i}[\log m_i - \log m(< x_i;c)]^2, 
\end{equation}
%
where $m(<x_i;c)$ corresponds to the values in Eq.(\ref{eq:profile}) at
$x=x_i$ for a given value of the concentration parameter $c$.

We sample the likelihood function distribution defined by ${\cal
  L}(c)=\exp(-\chi^2(c)/2)$ with a Metropolos-Hastings algorithms to
find the optimum value of $c$ and its  associated uncertainty
$\sigma_c$.

{\bf Cuantos pasos usa la cadena? De cuanto es el sigma en
los saltos de c en lca cadena?

Falta alguna figura que muestre resultados de la -log likelihood para
algunos casos representativos.
}

\section{Halo Samples}

We apply our method to two different halo samples. The first one is a
mock sample to test the algorithm and asses the impact of the number
of computational particles on the concentration values. The second
sample comes from a publicly available N-body cosmological
simulation. These results for these sample can be then compared
against other methods to find the concentration and feed the
discussion about the implications of our method.

@@

First, we test our method using mock halos with known values for the
concentration. The method we use to generate the halos is based on the integrated
mass profile. Given the number of particles $n$ and the concentration
$c$ we define the mass element as $\delta m = 1/n$ (this corresponds
to the mass of each particle such that the total mass is one). Then
for each number $k$ from $1$ to $n$ we find the value of $r$ such that
the difference 
%
\begin{equation}
m(<r;c)-k \cdot \delta m 
\end{equation}
%
is zero using Ridders' method. This value of $r$ is the radius of
the $k$th particle of the generated halo, then polar and azimuthal
angles $\theta$ and $\phi$ are randomly generated. Finally these three
coordinates are transformed into cartesian coordinates
$(r,\theta,\phi) \rightarrow (x,y,z)$. This process is repeated $n$
times in order to generate the $x,y,z$ coordinates for each particle.

We generate $100$ mock halos with concentration values randomly
placed in the range $1<c<20$. Each one of the halos is generated with
four different total particle numbers: $20$, $200$, $2000$ and
$20000$. This gives us a total of $500$ halos.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.48\textwidth]{error.pdf}
\end{center}
\caption{Relative error against number of particles
    \label{fig:error}}
\end{figure}

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.48\textwidth]{concentration.pdf}
\end{center}
\caption{Concentration against number of particles
    \label{fig:concentrations}}
\end{figure}
